---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clinfun

<!-- badges: start -->
[![R build status](https://github.com/karissawhiting/clinfun/workflows/R-CMD-check/badge.svg)](https://github.com/karissawhiting/clinfun/actions)
<!-- badges: end -->

Utilities to make your clinical collaborations easier if not fun. It contains functions for designing studies such as Simon 2-stage and group sequential designs and for data analysis such as Jonckheere-Terpstra test and estimating survival quantiles.

## Installation

You can install the released version of clinfun from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("clinfun")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("veseshan/clinfun")
```
## Example

Here is a basic example which shows how calculate CPS and Fisher's exact sample sizes with power estimates for given response rates of standard and experiment groups.

```{r example}
library(clinfun)

fe.ssize(p1 = .8,
         p2 = .2,
         alpha=0.05,
         power=0.8)
```

